---
title: "Effect of JQ1 and ingenol on human T cells"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(plater)
library(stringr)

# read in flow data ------------------------------------------------------------
# use readr::read_tsv because here is a weird NA symbol later in the table and 
# this way it gets turned into an NA instead of making all the columns factors
d <- read_tsv("../002 LRA/data/002 LRA gating-Table", na = "¥")

# better column names
colnames(d)[2:3] <- c("Well", "Plate")

# sample column is useless
d <- select(d, -Sample)

# read in plate layout data ----------------------------------------------------
dir <- "../002 LRA/plate_layout_"
mapping_files <- paste0(dir, c("combo_and_cntls.csv", "single_FD.csv", 
  "single_np.csv"))

# plate_mappings_combo has some character concentrations where the others
# have only numeric concentrations, which doesn't work with read_plates / 
# bind_rows, so have to do separately and use rbind
plate_mappings <- read_plates(mapping_files[2:3], plate_names = 
    c("free_drug_single", "nanoparticle_single"), well_ids_column = "Well")

plate_mappings_combo <- read_plate(mapping_files[1], 
    well_ids_column = "Well") %>% 
  mutate(Plate = "fd_combo_np_combo_cntls_np_only")

plate_mappings <- rbind(plate_mappings, plate_mappings_combo)

# data is ready ----------------------------------------------------------------
d <- inner_join(d, plate_mappings)
  
to_plot <- d %>% 
  # for combos just use concentration of one drug
  mutate(nM = as.numeric(str_replace(Concentration_nM, " .*", ""))) %>% 
  # omit nanoparticle alone
  filter(Donor != "Nanoparticles alone") %>% 
  # omit PMA
  filter(Drug != "PMA") %>%
  mutate(Donor = paste("Donor", Donor))

dmso <- filter(to_plot, Drug == "DMSO") %>% 
  group_by(Plate, Donor) %>% 
  summarise_at(vars(LivePercent:CD8HLADR), funs(mean))

tests <- to_plot %>% 
  filter(Drug != "DMSO") %>% 
  separate(Drug, into = c("Drug", "Formulation"), sep = "_") %>% 
  mutate(Drug = factor(Drug, levels = c("JQ1", "Ingenol", "both"))) %>% 
  mutate(Formulation = ifelse(Formulation == "FD", 
    "free drug", "nanoparticle")) 
```

## Goal

Determine the effect of JQ1 and Ingenol on human PBMC, using cellular gene expression as proxies for their ability to reactivate latent HIV.

## Experiment

24h exposure of human PBMC (two donors, 200K per well) to JQ1, Ingenol, or both, formulated as either free drugs or in nanoparticles. As a negative control, DMSO was used. After exposure, cells were stained for viability, cell lineage, and for markers that have been reported to respond to treatment with the drugs.

Drugs were diluted in threefold steps, typically starting at ninefold above Shijie's experiments. 

## Viability and cell number

We first assessed the viability of all cells, which is shown below. 

```{r}
helper <- function(y) {
  ggplot(tests, aes_string(x = "log10(nM)", y = y, color = "Formulation")) + 
    geom_hline(data = dmso, aes_string(yintercept = y)) +
    geom_point(size = 2) + facet_grid(Donor ~ Drug) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Formulation))
}

helper("LivePercent") + ylab("Viability") + ylim(c(0, 100))
```

Each symbol represents a single well, with colors representing drug formulations. The horizontal lines show the viability of the DMSO-treated wells. The concentration shown for "both" was the concentration of the JQ1 in the mixture. 

Only JQ1 as a free drug had no effect on the viability of the cells. All of the nanoparticles, as well as ingenol as a free drug, all had strong negative effects on the viability of the cells at higher concentrations. 

Similarly for cell number: 

```{r}
helper("log10(CD4Count)") + ylab("# Live CD4 cells (log10)") + ylim(c(1, 4))
helper("log10(CD8Count)") + ylab("# Live CD8 cells (log10)") + ylim(c(1, 4))
```

In order to avoid over-interpretation of wells where there were few, mostly-dead cells, we removed measurements from wells where overall viability was below 50% or CD4 or CD8 cell counts were below 500. This was typically the highest one or two concentration steps. 

```{r}
tests <- tests %>% 
  filter(LivePercent > 50, CD4Count > 500, CD8Count > 500)
```

## Gene proxies for JQ1

### CD28

JQ1 has been reported to down-regulate CD28 and CXCR4. 

We first tested the expression of CD28 on CD4+ T cells: 

```{r}
helper("CD4CD28") + ylab("CD28% of CD4 T cells") + ylim(c(0, 100))
```

... and CD8+ T cells ...

```{r}
helper("CD8CD28") + ylab("CD28% of CD8 T cells") + ylim(c(0, 100))
```

Almost all CD4 T cells were CD28+ to begin with and the drugs didn't seem to affect its expression. CD28 expression was more variable across donors on CD8 T cells. At the highest concentrations tested, it was up-regulated by JQ1 free drug. This effect was not present in the nanoparticle formulation or the combination with ingenol due to the toxicity of the nanoparticle formulation / ingenol at the highest concentration. 

### CXCR4

We next tested the expression of CXCR4 on CD4+ T cells: 

```{r}
helper("CD4CXCR4") + ylab("CXCR4% of CD4 T cells") + ylim(c(0, 60))
```

... and CD8+ T cells ...

```{r}
helper("CD8CXCR4") + ylab("CXCR4% of CD8 T cells") + ylim(c(0, 60))
```

Both JQ1 and ingenol caused downregulation of CXCR4. The downregulation occurred over a much narrower concentration window as well as to a larger extent for ingenol than for JQ1. The nanoparticle and free drug formulations were approximately equivalent. 

## Gene proxies for ingenol

### CD69

JQ1 has been reported to up-regulate CD69 and HLA-DR. 

We first tested the expression of CD69 on CD4+ T cells: 

```{r}
helper("CD4CD69") + ylab("CD69% of CD4 T cells") + ylim(c(0, 100))
```

... and CD8+ T cells ...

```{r}
helper("CD8CD69") + ylab("CD69% of CD8 T cells") + ylim(c(0, 100))
```

CD69 was dramatically upregulated on both CD4 and CD8 T cells by ingenol over a narrow concentration window. JQ1 had little to no effect. Nanoparticle and free drug had similar effects. 

### HLADR

We next tested the expression of HLADR on CD4+ T cells: 

```{r}
helper("CD4HLADR") + ylab("HLADR% of CD4 T cells") + ylim(c(0, 50))
```

... and CD8+ T cells ...

```{r}
helper("CD8HLADR") + ylab("HLADR% of CD8 T cells") + ylim(c(0, 50))
```

Ingenol caused upregulation of HLA-DR on CD8 T cells. It also caused a small upregulation on CD4 T cells, but to a much smaller extent. JQ1 had basically no effect. 

## Conclusions

### Proxies for LRA

CD28 is supposed to be downregulated by JQ1, but we found that it was unchanged (CD4 T cells) or upregulated (CD8 T cells). It was unaffected by ingenol.

CXCR4 is supposed to be downregulated by JQ1. We found that it was weakly downregulated by JQ1 and strongly downregulated by ingenol on both CD4 and CD8 T cells. 

CD69 is supposed to be upregulated by ingenol. We found that it was on both cell types. JQ1 had no effect.

HLADR is supposed to be upregulated by ingenol. It was slightly upregulated on CD4 T cells and moderately upregulated on CD8 T cells. JQ1 had no effect.  

**Best proxy for JQ1**: Hard to choose, none are very good. CXCR4 is "best" but not "good". CD40L, not tested here, has also been reported to be affected by JQ1.

**Best proxy for Ingenol**: CXCR5 and CD69 are both good: large effects and similar magnitudes between CD4 and CD8 T cells (which will help discriminating with the antibody-targeted particles)

### Little room for error

All drugs, except for JQ1 as a free drug, had large effects on viability, with only a narrow window (3- to 9-fold) between large toxic effect and maximum effect on protein expression. The nanoparticles by themselves seem to have a toxic effect, because cell viability slumped at high levels of JQ1 nanoparticle but not JQ1 free drug. 

### Nanoparticles vs. free drugs

The effect of drugs was broadly similar across formulations. CXCR4 was downregulated at lower concentrations (3- to 9-fold) by nanoparticle than by free drug. CD69 and HLADR were almost completely overlapping between the formulations. 