---
title: "Effect of JQ1 and Ingenol on human T cells"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(plater)
library(stringr)

# read in flow data ------------------------------------------------------------
# use readr::read_tsv because here is a weird NA symbol later in the table and 
# this way it gets turned into an NA instead of making all the columns factors
d <- read_tsv("../002 LRA/data/002 LRA gating-Table", na = "¥")

# better column names
colnames(d)[2:3] <- c("Well", "Plate")

# sample column is useless
d <- select(d, -Sample)

# read in plate layout data ----------------------------------------------------
dir <- "../002 LRA/plate_layout_"
mapping_files <- paste0(dir, c("combo_and_cntls.csv", "single_FD.csv", 
  "single_np.csv"))

# plate_mappings_combo has some character concentrations where the others
# have only numeric concentrations, which doesn't work with read_plates / 
# bind_rows, so have to do separately and use rbind
plate_mappings <- read_plates(mapping_files[2:3], plate_names = 
    c("free_drug_single", "nanoparticle_single"), well_ids_column = "Well")

plate_mappings_combo <- read_plate(mapping_files[1], 
    well_ids_column = "Well") %>% 
  mutate(Plate = "fd_combo_np_combo_cntls_np_only")

plate_mappings <- rbind(plate_mappings, plate_mappings_combo)

# data is ready ----------------------------------------------------------------
d <- inner_join(d, plate_mappings)
  
to_plot <- d %>% 
  # for combos just use concentration of one drug
  mutate(NgPerMl = as.numeric(str_replace(Concentration_ng_per_mL, " .*", ""))) %>% 
  # omit nanoparticle alone
  filter(Donor != "Nanoparticles alone") %>% 
  # omit PMA
  filter(Drug != "PMA") %>%
  mutate(Donor = paste("Donor", Donor))

dmso <- filter(to_plot, Drug == "DMSO") %>% 
  group_by(Plate, Donor) %>% 
  summarise_at(vars(LivePercent:CD8HLADR), funs(mean))

tests <- to_plot %>% 
  filter(Drug != "DMSO") %>% 
  separate(Drug, into = c("Drug", "Formulation"), sep = "_") %>% 
  mutate(Drug = factor(Drug, levels = c("JQ1", "Ingenol", "both"))) %>% 
  mutate(Formulation = ifelse(Formulation == "FD", 
    "free drug", "nanoparticle")) 
```

## Goal

Determine the effect of JQ1 and Ingenol on human PBMC, using cellular gene expression as proxies for their ability to reactivate latent HIV.

## Analysis version

This is an updated analysis performed on September 12, 2017, with corrected drug concentrations for the nanoparticle formulations.

## Experiment

24h exposure of human PBMC (two donors, 150K per well) to JQ1, Ingenol, or both, formulated as either free drugs or in nanoparticles. The JQ1 nanoparticles were chemically conjugated and the Ingenol nanoparticles were physically encapsulted.  As a negative control, DMSO was used at a concentration matching that of the highest drug concentration. After exposure, cells were stained for viability, cell lineage, and for markers that have been reported to respond to treatment with the drugs.

Drugs were diluted in threefold steps, with the middle dilution being the same concentrations that Shijie when he tested the drugs in combination in Jlat cells. 

In order to minimize drug loss for the fast-release JQ1 nanoparticles, we resuspended those after the Ingenol nanoparticles just before adding cells.

We attempted to use PMA/Ionomycin as a positive control, with the concentrations used by Jiang et al (doi: [10.1097/QAD.0000000000000289](https://www.ncbi.nlm.nih.gov/pubmed/?term=24804860)) but the cells had very poor viability so the results are not shown here. These are higher concentrations than we usually use so we'll try our normal dose next time. 

## Viability and cell number

We first assessed the viability of all cells, which is shown below. 

Each symbol represents a single well (each condition was run in two wells), with colors representing drug formulations. The horizontal lines show the viability of the DMSO-treated wells. The concentration shown for "both" was the concentration of the JQ1 in the mixture. 

Only JQ1 as a free drug had no effect on the viability of the cells. All of the nanoparticles, as well as Ingenol as a free drug, all had strong negative effects on the viability of the cells at higher concentrations. 
```{r}
helper <- function(y) {
  ggplot(tests, aes_string(x = "log10(NgPerMl)", y = y, color = "Formulation")) + 
    geom_hline(data = dmso, aes_string(yintercept = y)) +
    geom_point(size = 2) + facet_grid(Donor ~ Drug) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Formulation)) + 
    xlab("ng/mL (log10)")
}

helper("LivePercent") + ylab("Viability") + ylim(c(0, 100))
```

Similarly for cell number: 

```{r}
helper("log10(CD4Count)") + ylab("# Live CD4 cells (log10)") + ylim(c(1, 4))
helper("log10(CD8Count)") + ylab("# Live CD8 cells (log10)") + ylim(c(1, 4))
```

In order to avoid over-interpretation of wells where there were few, mostly-dead cells, we removed measurements from wells where overall viability was below 50% or CD4 or CD8 cell counts were below 500. This was typically the highest one or two concentration steps. 

```{r}
tests <- tests %>% 
  filter(LivePercent > 50, CD4Count > 500, CD8Count > 500)
```

## Gene proxies for JQ1

### CD28

JQ1 has been reported to down-regulate CD28 and CXCR4 (Banerjee et al). 

We first tested the expression of CD28 on CD4+ T cells: 

```{r}
helper("CD4CD28") + ylab("CD28% of CD4 T cells") + ylim(c(0, 100))
```

... and CD8+ T cells ...

```{r}
helper("CD8CD28") + ylab("CD28% of CD8 T cells") + ylim(c(0, 100))
```

Almost all CD4 T cells were CD28+ to begin with and the drugs didn't seem to affect its expression. CD28 expression was more variable across donors on CD8 T cells. At the highest concentrations tested, it was up-regulated by JQ1 free drug. This effect was not present in the nanoparticle formulation or the combination with Ingenol due to the toxicity of the nanoparticle formulation / Ingenol at the highest concentration. 

### CXCR4

```{r}
helper("CD4CXCR4") + ylab("CXCR4% of CD4 T cells") + ylim(c(0, 60))
```

... and CD8+ T cells ...

```{r}
helper("CD8CXCR4") + ylab("CXCR4% of CD8 T cells") + ylim(c(0, 60))
```

Both JQ1 and Ingenol caused downregulation of CXCR4. The downregulation occurred over a much narrower concentration window as well as to a larger extent for Ingenol than for JQ1. The nanoparticle and free drug formulations were approximately equivalent for JQ1, while ingenol nanoparticles were 10-20X less potent than the free drug. 

## Gene proxies for Ingenol

### CD69

Ingenol alone and in combination with JQ1 has been reported to up-regulate CD69 and HLA-DR (Jiang et al and Darcis et al). 

We first tested the expression of CD69 on CD4+ T cells: 

```{r}
helper("CD4CD69") + ylab("CD69% of CD4 T cells") + ylim(c(0, 100))
```

... and CD8+ T cells ...

```{r}
helper("CD8CD69") + ylab("CD69% of CD8 T cells") + ylim(c(0, 100))
```

CD69 was dramatically upregulated on both CD4 and CD8 T cells by Ingenol over a narrow concentration window. JQ1 had little to no effect. Nanoparticle and free drug had similar effects, but again ingenol nanoparticles were 10-20X less potent. 

### HLADR

We next tested the expression of HLADR on CD4+ T cells: 

```{r}
helper("CD4HLADR") + ylab("HLADR% of CD4 T cells") + ylim(c(0, 50))
```

... and CD8+ T cells ...

```{r}
helper("CD8HLADR") + ylab("HLADR% of CD8 T cells") + ylim(c(0, 50))
```

Ingenol caused upregulation of HLA-DR on CD8 T cells as was reported in Jiang et al. It also caused a small upregulation on CD4 T cells, but to a much smaller extent. JQ1 had basically no effect. 

### Median fluorescence intensity of CD4 and CD8

PKC agonists reportedly reduce the expression on CD4. Combining them with JQ1 reportedly amplifies this effect. Therefore we checked how the drugs affected the MFI of CD4 and CD8. 

```{r}
helper("CD4MFI") + ylab("CD4 MFI")
helper("CD8MFI") + ylab("CD8 MFI")
```

Both JQ1 and ingenol reduced the expression of CD4, JQ1 to a much greater extent. At equivalent concentrations the downregulated was a bit more pronounced with the combination of the two. However, the maximum concentration of drug that could be reached without too much toxicity was lower for the combination. Both drugs also caused downregulation of CD8, but to a smaller extent, and again the combination was more powerful. 

The MFI of CD4 and CD8 was lower for the DMSO wells for unknown reasons. In all cases, drug-treated or not, positive CD4 and CD8 populations could be clearly distinguished. 

## Conclusions

### Proxies for LRA

CD28 is supposed to be downregulated by JQ1 (Bannerjee et al, in microarray and qPCR validated in Jlat 10.6 cells), but we found that it was unchanged (CD4 T cells) or upregulated (CD8 T cells). It was unaffected by Ingenol.

CXCR4 is supposed to be downregulated by JQ1 (Bannerjee et al, in microarray). We found that it was weakly downregulated by JQ1 and strongly downregulated by Ingenol on both CD4 and CD8 T cells. 

CD69 is supposed to be upregulated by Ingenol. We found that it was on both cell types. JQ1 had no effect.

HLADR is supposed to be upregulated by Ingenol. It was slightly upregulated on CD4 T cells and moderately upregulated on CD8 T cells. JQ1 had no effect.  

CD4, and to a lesser extent CD8, are also downregulated by both drugs. 

**Best proxy for JQ1**: Hard to choose, none are very good. CXCR4 is "best" but not "good". CD40L, not tested here, has also been reported to be affected by JQ1.

**Best proxy for Ingenol**: CXCR5 and CD69 are both good: large effects and similar magnitudes between CD4 and CD8 T cells (which will help discriminating with the antibody-targeted particles)

### Little room for error

All drugs, except for JQ1 as a free drug, had large effects on viability, with only a narrow window (3- to 9-fold) between large toxic effect and maximum effect on protein expression. The nanoparticles by themselves seem to have a toxic effect, because cell viability slumped at high levels of JQ1 nanoparticle but not JQ1 free drug. 

### Nanoparticles vs. free drugs

The effect of drugs was broadly similar across formulations. The ingenol nanoparticles required about 10-20X higher concentration to have the same effect as the free drug, while JQ1 nanoparticles and free drug had the same effect at about the same concentration. 