---
title: "Anti-proliferative effect of MPA-containing CD4 targeting nanoparticles (002_MPA)"
output: html_document
---

## Goal
Compare targeting of various formulations of MPA nanoparticles by measuring proliferation of CD8+ and CD4+ T cells. 

## Experiment

5 day exposure of anti-CD3/anti-CD28 bead stimulated and Cell Trace Violet stained human PBMC (two donors, 100k cells per well, in duplicate) to 6 nanoparticle formulations:



1.	MPA nanoparticles (un-targeted)

2.	MPA nanoparticles with BP4 (CD4-targeting peptide)

3.	MPA nanoparticles with CD4 antibody

4.	MPA nanoparticles with isotype antibody

5.	MPA nanoparticles with fragment CD4 antibody

5.	MPA nanoparticles with fragment isotype antibody




To control for the effect of nanoparticles we also exposed cells to MPA free drug. As a negative control, DMSO was used at a concentration matching that of the highest drug concentration. Formulations were diluted in two-fold steps.

Free MPA starting concentration: 0.4ug/mL

Nanoparticle formulations starting concentration: 8ug/mL (20x free drug)


After exposure, cells were stained for viability and  cell lineage.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plater)
library(stringr)
library(ggplot2)
```


```{r read in flow data}


dat <- read.csv("../002_MPA/Table.csv")


```



```{r}
# read in plate layout data ----------------------------------------------------
dir <- "../002_MPA/plate_layout_"
mapping_files <- paste0(dir, c("BP4_control_beads.csv",
                               "CD4_fragment_targeted_beads.csv",
                               "CD4_targeted_beads.csv",
                               "free_and_untargeted_beads.csv"))

#read in the plate layouts in and give them the same names
plate_mappings <- plater::read_plates(mapping_files, plate_names = 
   c("BP4_controls",
                               "fragment_isotype",
                               "CD4_isotype",
                               "bare_free") , well_ids_column = "Well")

```


```{r}

dat <- inner_join(dat, plate_mappings)


#remove the "nanoparticles without cells" wells, they were just there for setting gates.


dat<- dat %>%
  filter(Drug != "Nanoparticles without cells")


#the values in the Targeting column are long and won't fit well on my plots so I'll add some line breaks here.

dat <- dat %>%
  mutate(Targeting =  str_replace(Targeting, "Isotype for CD4 antibody", "Isotype for\n CD4 antibody"))%>%
  mutate(Targeting = str_replace(Targeting, "Fragment CD4 antibody", "Fragment\n CD4 antibody"))%>%
  mutate(Targeting = str_replace(Targeting,"Isotype for fragment CD4 antibody", "Isotype for\n CD4 fragment\n antibody") ) %>%
  mutate(Targeting = str_replace(Targeting,"Non-Targeted Nanoparticles", "Non-Targeted\n Nanoparticles"))

#also make the targeting values factors and put them in my preferred order!!


dat$Targeting <-factor(dat$Targeting, levels = c("CD4 antibody", "Isotype for\n CD4 antibody", "Fragment\n CD4 antibody", "Isotype for\n CD4 fragment\n antibody", "BP4", "Non-Targeted\n Nanoparticles","Free Drug" ))
```


```{r sanity check CD4}
#quick look at data: CD4 and CD8 and CD8Neg counts 



# ggplot(dat, aes(x = log10(NgPerMl), y = CD4Count))+
#   geom_point(aes(color = Formulation))+
#   facet_wrap(~Targeting)
# 
# ggplot(dat, aes(x = log10(NgPerMl), y = CD4CTVLow))+
#   geom_point(aes(color = Formulation))+
#   facet_wrap(~Targeting)
# ```
# 
# ```{r sanity check CD8}
# 
# ggplot(dat, aes(x = log10(NgPerMl), y = CD8Count))+
#   geom_point(aes(color = Formulation))+
#   facet_wrap(~Targeting)
# 
# ggplot(dat, aes(x = log10(NgPerMl), y = CD8CTVLow))+
#   geom_point(aes(color = Formulation))+
#   facet_wrap(~Targeting)
# ```
# 
# ```{r sanity check CD8Neg}
# 
# ggplot(dat, aes(x = log10(NgPerMl), y = CD8NegCount))+
#   geom_point(aes(color = Formulation))+
#   facet_wrap(~Targeting)
# 
# ggplot(dat, aes(x = log10(NgPerMl), y = CD8NegCTVLow))+
#   geom_point(aes(color = Formulation))+
#   facet_wrap(~Targeting)
```
## Viability


We first assessed the viability of all cells, which is shown below.

Each symbol represents a single well (each condition was run in two wells), with colors representing drug formulations. The horizontal lines show the viability of the DMSO-treated wells.

As expected, viability decreased as the concentration of MPA increased.

The points with high viability at higher concentrations represent cases where there were very few cells in the gate, but they happened to be live.



```{r create DMSO and nonDMSO dfs}

dmso <-  dat %>%
  filter(Drug == "DMSO")%>%
  group_by(Plate, Donor) %>% 
  summarise_at(vars(Live:CD8NegCTVLow), funs(mean))%>%
  ungroup()

no_DMSO<-filter(dat, Drug != "DMSO")



```

```{r plot viability}
ggplot(no_DMSO, aes(x = log10(NgPerMl), y = Live, color = Formulation)) + 
    geom_hline(data = dmso, aes(yintercept = Live)) +
    geom_point(size = 2) +
    facet_grid(~Donor) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Formulation)) + 
    xlab("ng/mL (log10)")
```

```{r filtering out low cell counts}

#want to keep data where either CD8Count >100 or CD8Neg>100, otherwise remove the row.

#To get CD8 high count data, select all cols EXCEPT CD8Neg ones. Filter for CD8Count >100. Gather. Repeat for CD8Neg, then rbind.

CD8_high_count <- no_DMSO %>%
  select(-CD8NegCount, -CD8Neg, -CD8NegCTVLow)%>%
  filter(CD8Count > 100)%>%
  gather(., key = Gate , value = Measurement,CD8:CD8CTVLow)
  

CD8Neg_high_count <- no_DMSO %>%
  select(-CD8Count, -CD8, -CD8CTVLow)%>%
  filter(CD8NegCount > 100)%>%
  gather(., key = Gate , value = Measurement,CD8Neg:CD8NegCTVLow)


high_count <- rbind(CD8Neg_high_count, CD8_high_count)
#Something weird happened with the staining in the fragment_isotype plate in wells F04 and F05. #Cells are 100% live and >90% of them are CD8Neg. In F04 we counted 1775 CD8Neg cells but only 285 in F05.  >90% prolif in both. I will exclude them from the analysis here.


high_count <-filter(high_count, X != "fragment_isotype F04" & X != "fragment_isotype F05")



```

## CD4 staining

As in the most recent LRA experiment with targeting nanoparticles, some of the targeting nanoparticles seem to interfere with the phenotyping stain for CD4+ cells, so we are using a CD8- gate to represent CD4+ cells in this analysis.

Data are not shown for wells where cell counts were <100.


```{r CD4 staining}

high_count %>%
  filter(Gate == "CD8"|Gate == "CD8Neg" |Gate == "CD4")%>%
ggplot(., aes(x = log10(NgPerMl), Measurement )) +
  geom_point(aes(color = Gate))+
  stat_summary(fun.y = "mean", geom = "line", aes(group = Gate, color = Gate)) +
  facet_grid(Donor~Targeting)+
  labs(x = "ng/mL (log10)", y = "Percent in gate")+
  theme(strip.text = element_text(size = 8))

 

```


```{r gather dmso data}

#make the dmso data long for plotting proliferation.
prolif_dmso <-dmso %>%
  select(Plate, Donor,CD8CTVLow, CD8NegCTVLow)%>%
  gather(., key = Gate, value = Measurement, CD8CTVLow, CD8NegCTVLow)


```

```{r prolif plot function}
#here is a function for making the plots for the percent prolif results

prolif_plot <- function(df_to_plot){
  ggplot(df_to_plot, aes(x = log10(NgPerMl), y = Measurement, color = Gate)) + 
geom_hline(data = prolif_dmso, aes(yintercept = Measurement, color = Gate))+
geom_point(size = 2) +
facet_grid(Donor~Targeting) + 
stat_summary(fun.y = "mean", geom = "line", aes(group = Gate)) + 
labs(x = "ng/mL (log10)", y = "Percent Proliferated")
}

```


```{r proliferation data}


high_count_prolif <- high_count %>%
  filter(Gate == "CD8NegCTVLow" | Gate == "CD8CTVLow")



```

## Anti-proliferative results

Data are not shown for wells with cells counts <100. Horizontal lines show results for DMSO controls.

### anti-CD4 antibody targeting and isotype control


```{r CD4 targeting prolif}
#Filter for CD8CTVLow and CD8NegCTVLow and for the antiCD4- ab targeting

cd4_cd4_iso_prolif <- high_count_prolif%>%
  filter(Targeting == "CD4 antibody"| Targeting == "Isotype for\n CD4 antibody" | Targeting == "Non-Targeted\n Nanoparticles"|Targeting == "Free Drug")


prolif_plot(cd4_cd4_iso_prolif)

```




### Fragment anti-CD4 antibody targeting and isotype control

```{r frament and iso}

fragment_and_fiso_prolif <- high_count_prolif%>%
  filter(Targeting == "Fragment\n CD4 antibody"| Targeting == "Isotype for\n CD4 fragment\n antibody"| Targeting == "Non-Targeted\n Nanoparticles"|Targeting == "Free Drug")
  
  

prolif_plot(fragment_and_fiso_prolif)
```


### BP4 targeting

```{r bp4}

bp4_prolif <-high_count_prolif%>%
  filter(Targeting == "BP4"|Targeting == "Non-Targeted\n Nanoparticles"|Targeting == "Free Drug")
  
prolif_plot(bp4_prolif)

```

### Proliferation ratios


The following plot shows the ratio of the CD8+ CTV Low percentage over the CD8- CTV Low percentage. The horizontal line is at 1, where the percent proliferation is equal between the two cell types. 
*Above* the blue line means that CD8+ proliferated more than CD8- cells.

```{r CTV ratios}

# to make a column for plotting of CD8NegCTVLow/CD8CTVLow, first spread the data back into wide format, then make a column for the ratio
wide_high_count_prolif <- high_count_prolif %>%
  spread(., Gate, Measurement)%>%
  mutate(CD8_CD8Neg_ratio = CD8CTVLow/CD8NegCTVLow)
  
  ggplot(wide_high_count_prolif, aes(x = log10(NgPerMl), CD8_CD8Neg_ratio )) +
  geom_point(aes())+
  stat_summary(fun.y = "mean", geom = "line", aes()) +
  facet_grid(Donor~Targeting)+
  labs(x = "ng/mL (log10)", y = "CD8+/CD8- Proliferation")+
  geom_hline(yintercept = 1, color = "blue")+
  theme(strip.text = element_text(size = 8))
```


## Conclusions

* A comparison of the Fragment CD4 antibody and its isotype control shows some targeting. 

* Difficult to disentangle an anti-proliferative effect from cell death.

* Viability wasn't great for this experiment even in our DMSO control, so could repeat with different donors. 

## Questions

* What is the effect of nanoparticles that aren't bound to CD4+ cells releasing drug?

* Could we expose cells to nanoparticles and wash off un-bound particles?


