---
title: "Effect of Ingenol-containing CD4 targeting nanoparticle formulations on CD69 (003 LRA)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(plater)
library(stringr)
```


```{r read in flow data}
# read in flow data ------------------------------------------------------------
# use readr::read_tsv because here is a weird NA symbol later in the table and 
# this way it gets turned into an NA instead of making all the columns factors
d <- read.csv("../003_LRA/14-Sep-2017 FlowJo table.csv")


#the "." in the colnames should be a "-" sign but that is annoying so I'll replace with "Neg"

colnames(d)<- str_replace(colnames(d),"\\.","Neg")


# read in plate layout data ----------------------------------------------------
dir <- "../003_LRA/plate_layout_"
mapping_files <- paste0(dir, c("BP4_and_control.csv",
                               "CD4_and_CD4_iso.csv",
                               "fCD4_and_fCD4_iso.csv",
                               "free_and_bare.csv"))

#read in the plate layouts in and give them the same names
plate_mappings <- read_plates(mapping_files, plate_names = 
   c("BP4_and_control",
                               "CD4_and_CD4_iso",
                               "fCD4_and_fCD4_iso",
                               "free_and_bare") , well_ids_column = "Well")

```


```{r}
# data is ready ----------------------------------------------------------------
d <- inner_join(d, plate_mappings)
  
to_plot <- d %>% 
  # for combos just use concentration of one drug
  mutate(NgPerMl = as.numeric(str_replace(Concentration_ng_per_mL, " .*", ""))) %>% 
  # omit nanoparticle alone
  filter(Donor != "Nanoparticles without cells") %>% 
  mutate(Donor = paste("Donor", Donor))

dmso <- filter(to_plot, Drug == "DMSO") %>% 
  group_by(Plate, Donor) %>% 
  summarise_at(vars(Live:CD8NegCD69MFI), funs(mean))

tests <- to_plot %>% 
  filter(Drug != "DMSO") %>% 
  mutate(Formulation = ifelse(Formulation == "Free", 
    "Free Drug", "Nanoparticle")) 
```

## Goal

Use the proxy marker CD69 to test the effectivness of various CD4+ cell targeting nanoparticle formulations that are chemically conjugated to the LRA Ingenol. A previous experiment (002_LRA) showed that Ingenol up-regulated CD69 in both CD4+ and CD8+ T cells. If the targeting is effective, we expect to repeat up-regulation of CD69 in CD4+ T cells but not in (un-targeted) CD8+ T cells.


## Experiment

24h exposure of human PBMC (two donors, 100k-150k cells per well, in duplicate) to 6 nanoparticle formulations:


1.	Ingenol nanoparticles

2.	Ingenol nanoparticles with BP4 (CD4-targeting peptide)

3.	Ingenol nanoparticles with CD4 antibody

4.	Ingenol nanoparticles with isotype antibody

5.	Ingenol nanoparticles with fragment CD4 antibody

6.	Ingenol nanoparticles with fragment isotype antibody


To control for the effect of nanoparticles we also exposed cells to Ingenol free drug. The nanoparticle Ingenol concentration was 20x that of the free drug.  *Shijie's experiments showed that you should use ~50X dose for nanoparticles but we didn't have enough nanoparticles so had to do 20X*

As a negative control, DMSO was used at a concentration matching that of the highest drug concentration. 

After exposure, cells were stained for viability, cell lineage, and for CD69, which we showed to be up-regulated in both CD4+ and CD8+ T cells upon exposure to Ingenol in a previous experiment.

Drugs were diluted in threefold steps, with the middle dilution being the same concentrations that Shijie when he tested the drugs in combination in Jlat cells. 



## Viability and cell number

We first assessed the viability of all cells, which is shown below. 

Each symbol represents a single well (each condition was run in two wells), with colors representing drug formulations. The horizontal lines show the viability of the DMSO-treated wells. 

 
```{r}
helper <- function(y) {
  ggplot(tests, aes_string(x = "log10(NgPerMl)", y = y, color = "Formulation")) + 
    geom_hline(data = dmso, aes_string(yintercept = y)) +
    geom_point(size = 2) +
    facet_grid(~Donor) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Formulation)) + 
    xlab("ng/mL (log10)")
}

helper("Live") + ylab("Viability") + ylim(c(0, 100))
```

## Targeting nanoparticles interfere with CD4+ staining

Flow plots showed diminished CD4+ populations at higher concentrations of nanoparticles. In contrast, gating on CD8- to capture CD4+ cells did not show this pattern. This suggests that the nanoparticles may have interfered with binding of the anti-CD4 antibody that we used for phenotyping the cells.

Because of this effect, we used the percentages from the CD8- gates, rather than the CD4+ gates, for the rest of the analyses for this experiment.


```{r CD4 staining}
#I want to show that CD4 staining didn't work very well, especially with higher drug concentrations and that CD8Neg is a better way to look at the CD4+ cells.

# select just the meta data columns and the CD4, CD8 and CD8Neg.
#Gather the data so all the Percentages are in a column called Percent and all the Gates (CD4, CD8 and CD8Neg) that the percentages belong to are in another column.

# Then plot the NgPerMl vs the Percent in the different gates (make gates different shapes) and facet by donor.

CD4_vs_CD8Neg_staining <- tests %>%
  select(CD4, CD8, CD8Neg, Formulation, Targeting, NgPerMl, Donor)%>%
  gather(data = ., key = Gate, value = Percent, CD4, CD8, CD8Neg)

ggplot(CD4_vs_CD8Neg_staining, aes(x = log10(NgPerMl), Percent )) +
  geom_point(aes(color = Gate))+
  facet_grid(Donor~Formulation)

```


## Percent CD69+ 

### anti-CD4 antibody targeting and isotype control

#### CD8- T cells


```{r}

#Want to look at the percentages in the CD8CD69 gates vs. the CD8NegCD69 gates for each of the nanoparticle formulations.

#If targeting worked, should see higher percentages in the CD8NegCD69 gates with the targeted NPs and about the same for the isotype controls and for free drug. 

tests %>%
  filter(Targeting == "CD4 antibody"| Targeting == "Isotype for CD4 antibody" | Targeting == "None"|Targeting == "Free Drug")%>%
ggplot(aes_string(x = "log10(NgPerMl)", y = "CD8NegCD69", color = "Targeting")) + 
    geom_hline(data = dmso, aes_string(yintercept = "CD8NegCD69"))+
  geom_point(size = 2) +
    facet_grid(~Donor) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Targeting)) + 
    xlab("ng/mL (log10)")


```

### CD8+ T cells

```{r}

tests %>%
  filter(Targeting == "CD4 antibody"| Targeting == "Isotype for CD4 antibody" | Targeting == "None"|Targeting == "Free Drug")%>%
ggplot(aes_string(x = "log10(NgPerMl)", y = "CD8CD69", color = "Targeting")) + 
    geom_hline(data = dmso, aes_string(yintercept = "CD8CD69"))+
  geom_point(size = 2) +
    facet_grid(~Donor) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Targeting)) + 
    xlab("ng/mL (log10)")






#want one plot per targeting type + isotype control, faceted by donor and by gate type (gates are CD8NegCD69 and CD8CD69).


# In order to facet by gate, need to gather the data from the CD8NegCD69 and CD8CD69, like I did in the CD4 staining chunk above.

#I'll add the hline for the dmso CD69, but I'll have to modify the dmso df to make it work with my plot



dmso_CD69_results <-dmso %>%
  select(CD8CD69, CD8NegCD69,Donor)%>%
  gather(data = ., key = Gate, value = Percent,CD8CD69, CD8NegCD69)

CD69_results <- tests %>%
  select(CD8CD69, CD8NegCD69, Formulation, Targeting, NgPerMl, Donor,Drug)%>%
  gather(data = ., key = Gate, value = Percent,CD8CD69, CD8NegCD69)


CD69_results%>%
  filter(Targeting == "CD4 antibody"| Targeting == "Isotype for CD4 antibody" | Targeting == "None"|Targeting == "Free Drug") %>%
  
  ggplot(aes_string(x = "log10(NgPerMl)", y = "Percent", color = "Targeting")) + 
    geom_hline(aes(yintercept = Percent),data = dmso_CD69_results)+
  geom_point(aes(shape = Drug),size = 2) +
    facet_grid(Donor~Gate) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Targeting)) + 
    xlab("ng/mL (log10)")
  




```
Similarly for cell number: 

```{r}
# helper("log10(CD4Count)") + ylab("# Live CD4 cells (log10)") + ylim(c(1, 4))
# helper("log10(CD8Count)") + ylab("# Live CD8 cells (log10)") + ylim(c(1, 4))
```

In order to avoid over-interpretation of wells where there were few, mostly-dead cells, we removed measurements from wells where overall viability was below 50% or CD4 or CD8 cell counts were below 500. This was typically the highest one or two concentration steps. 

```{r}
# tests <- tests %>% 
  # filter(Live > 50, CD4Count > 500, CD8Count > 500)
```


## Gene proxies for Ingenol

### CD69



```{r}
# helper("CD4CD69") + ylab("CD69% of CD4 T cells") + ylim(c(0, 100))
```

... and CD8+ T cells ...

```{r}
# helper("CD8CD69") + ylab("CD69% of CD8 T cells") + ylim(c(0, 100))
```


### Median fluorescence intensity of CD4 and CD8



```{r}

```



## Conclusions

