---
title: "Effect of Ingenol-containing CD4 targeting nanoparticle formulations on CD69 (003 LRA)"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plater)
library(stringr)
```


```{r read in flow data}
# read in flow data ------------------------------------------------------------
# use readr::read_tsv because here is a weird NA symbol later in the table and 
# this way it gets turned into an NA instead of making all the columns factors
d <- read.csv("../003_LRA/14-Sep-2017 FlowJo table.csv")


#the "." in the colnames should be a "-" sign but that is annoying so I'll replace with "Neg"

colnames(d)<- str_replace(colnames(d),"\\.","Neg")
```



```{r}
# read in plate layout data ----------------------------------------------------
dir <- "../003_LRA/plate_layout_"
mapping_files <- paste0(dir, c("BP4_and_control.csv",
                               "CD4_and_CD4_iso.csv",
                               "fCD4_and_fCD4_iso.csv",
                               "free_and_bare.csv"))

#read in the plate layouts in and give them the same names
plate_mappings <- plater::read_plates(mapping_files, plate_names = 
   c("BP4_and_control",
                               "CD4_and_CD4_iso",
                               "fCD4_and_fCD4_iso",
                               "free_and_bare") , well_ids_column = "Well")

```


```{r}
# data is ready ----------------------------------------------------------------
d <- inner_join(d, plate_mappings)
  

#In addition to removing the rows where the Drug is "Nanoparticles without cells" I will also remove the CD4CD69MFI column because I'm not going to use it (since the CD4 staining had problems I'll use CD8- to represent CD4+ cells). since we counted 0% CD4 cells in that well, the CD4CD69MFI has "n/a" which screws up my classes and prevents the MFI values from being numeric.


to_plot <- d %>% 
  # for combos just use concentration of one drug
  mutate(NgPerMl = as.numeric(str_replace(Concentration_ng_per_mL, " .*", ""))) %>% 
  # omit nanoparticle alone
  filter(Drug != "Nanoparticles without cells") %>% 
  mutate(Donor = paste("Donor", Donor))%>%
  select(-CD4CD69MFI)


# When there are no cells in a gate, the MFI is given as n/a. This was the case in the "Nanoparticles without cells" wells in the "d" dataframe and in one well where the CD4 staining apparently didn't work (fCD4_and_fCD4_iso, C03). 

#This caused the class of the columns that had n/a (CD4CD69MFI and CD8CD69MFI)to become factors.I fixed the problem wit the CD4CD69MFI problem by removing it, but after filtering out those ""Nanoparticles , the incorrect type carried over in the CD8CD69MFI column so I need to change it here.  

#Then I'll convert the remaining problematic column, CD8CD69MFI to numeric (After first converting to character to avoid get the number of the factor level!!!!)

to_plot$CD8CD69MFI <- as.numeric(as.character(to_plot$CD8CD69MFI))

```


```{r make dmso and no dmso dfs}
dmso <- filter(to_plot, Drug == "DMSO") %>% 
  group_by(Plate, Donor) %>% 
  summarise_at(vars(Live:CD8NegCD69MFI), funs(mean))%>%
  ungroup()

tests <- to_plot %>% 
  filter(Drug != "DMSO") %>% 
  mutate(Formulation = ifelse(Formulation == "Free", 
    "Free Drug", "Nanoparticle")) %>%
  mutate(Targeting = str_replace(Targeting, "None", "Non-Targeted\nNanoparticles"))


#make targeting a factor and order in the way I want on plots

tests$Targeting <- factor(tests$Target, levels = c( "CD4 antibody","Isotype for CD4 antibody","Fragment CD4 antibody", "Isotype for fragment CD4 antibody", "BP4","Free Drug","Non-Targeted\nNanoparticles" ))


#the values in the Targeting column are long and won't fit well on my plots so I'll add some line breaks here.

tests <- tests %>%
  mutate(Targeting =  str_replace(Targeting, "Isotype for CD4 antibody", "Isotype for\nCD4 antibody"))%>%
  mutate(Targeting = str_replace(Targeting, "Fragment CD4 antibody", "Fragment\nCD4 antibody"))%>%
  mutate(Targeting = str_replace(Targeting,"Isotype for fragment CD4 antibody", "Isotype for\nCD4 fragment\nantibody") )


```

## Goal

Use the proxy marker CD69 to test the effectivness of various CD4+ cell targeting nanoparticle formulations that are chemically conjugated to the LRA Ingenol. A previous experiment (002_LRA) showed that Ingenol up-regulated CD69 in both CD4+ and CD8+ T cells. If the targeting is effective, we expect to repeat up-regulation of CD69 in CD4+ T cells but not in (un-targeted) CD8+ T cells.


## Experiment

24h exposure of human PBMC (two donors, 100k-150k cells per well, in duplicate) to 6 nanoparticle formulations:


1.	Ingenol nanoparticles

2.	Ingenol nanoparticles with BP4 (CD4-targeting peptide)

3.	Ingenol nanoparticles with CD4 antibody

4.	Ingenol nanoparticles with isotype antibody

5.	Ingenol nanoparticles with fragment CD4 antibody

6.	Ingenol nanoparticles with fragment isotype antibody


To control for the effect of nanoparticles we also exposed cells to Ingenol free drug. The nanoparticle Ingenol concentration was 20x that of the free drug.  *Shijie's experiments showed that you should use ~50X dose for nanoparticles but we didn't have enough nanoparticles so had to do 20X*

As a negative control, DMSO was used at a concentration matching that of the highest drug concentration. 

After exposure, cells were stained for viability, cell lineage, and for CD69, which we showed to be up-regulated in both CD4+ and CD8+ T cells upon exposure to Ingenol in a previous experiment.

Drugs were diluted in threefold steps, with the middle dilution being the same concentrations that Shijie used when he tested the drugs in combination in Jlat cells. 



## Viability and cell number

We first assessed the viability of all cells, which is shown below. 

Each symbol represents a single well (each condition was run in two wells), with colors representing drug formulations. The horizontal lines show the viability of the DMSO-treated wells. 


Viability decreased as the concentration of Ingenol increased, but not as dramactically as in the last experiment.
 
```{r viability}
#this is a helper function for plotting.

helper <- function(y) {
  ggplot(tests, aes_string(x = "log10(NgPerMl)", y = y, color = "Formulation")) + 
    geom_hline(data = dmso, aes_string(yintercept = y)) +
    geom_point(size = 2) +
    facet_grid(~Donor) + 
    stat_summary(fun.y = "mean", geom = "line", aes(group = Formulation)) + 
    xlab("ng/mL (log10)")
}

helper("Live") + ylab("Viability") + ylim(c(0, 100))

```



## Targeting nanoparticles interfere with CD4+ staining

Flow plots showed diminished CD4+ populations at higher concentrations of nanoparticles. In contrast, gating on CD8- to capture CD4+ cells did not show this pattern. This suggests that the nanoparticles may have interfered with binding of the anti-CD4 antibody that we used for phenotyping the cells. The effect is strongest for the targeting nanoparticles but is also present in the isotype controls and non-targeted nanoparticles.

This is shown in the plot below, where the red dots, representing the percentage of cells in the CD4+ gate drop to lower percentages as the drug concentration increases along the x-axis.

Because of this effect, we used the percentages from the CD8- gates, rather than the CD4+ gates, for the rest of the analyses for this experiment.



```{r CD4 staining}
#I want to show that CD4 staining didn't work very well, especially with higher drug concentrations and that CD8Neg is a better way to look at the CD4+ cells.

# select just the meta data columns and the CD4, CD8 and CD8Neg.
#Gather the data so all the Percentages are in a column called Percent and all the Gates (CD4, CD8 and CD8Neg) that the percentages belong to are in another column.

# Then plot the NgPerMl vs the Percent in the different gates (make gates different shapes) and facet by donor.

CD4_vs_CD8Neg_staining <- tests %>%
  select(CD4, CD8, CD8Neg, Formulation, Targeting, NgPerMl, Donor)%>%
  gather(data = ., key = Gate, value = Percent, CD4, CD8, CD8Neg)

ggplot(CD4_vs_CD8Neg_staining, aes(x = log10(NgPerMl), Percent )) +
  geom_point(aes(color = Gate))+
  stat_summary(fun.y = "mean", geom = "line", aes(group = Gate, color = Gate)) +
  facet_grid(Donor~Targeting)+
  labs(x = "ng/mL (log10)", y = "Percent in gate")+
  theme(strip.text = element_text(size = 8))

```


## Percent CD69+


Symbols represent experimental replicates and the connecting lines show the average of the replicates. The horizontal lines represent the DMSO negative controls.

```{r}

#want one plot per targeting type + isotype control, faceted by donor and by gate type (gates are CD8NegCD69 and CD8CD69).


# In order to facet by gate, need to gather the data from the CD8NegCD69 and CD8CD69, like I did in the CD4 staining chunk above.




CD69_percent <- tests %>%
  select(CD8CD69, CD8NegCD69, Formulation, Targeting, NgPerMl, Donor,Drug)%>%
  gather(data = ., key = Gate, value = Percent,CD8CD69, CD8NegCD69)






#I'll add the hline for the dmso CD69, but I'll have to modify the dmso df to make it work with my plot
dmso_CD69_percent <-dmso %>%
  select(CD8CD69, CD8NegCD69,Donor)%>%
  gather(data = ., key = Gate, value = Percent,CD8CD69, CD8NegCD69)
```


```{r percent plot functions}
#here is a function for making the plots for the percent CD69 results


percent_plot <-function(df_to_plot){
  
ggplot(df_to_plot, aes(x = log10(NgPerMl), y = Percent, color = Targeting)) + 
geom_hline(aes(yintercept = Percent),
           data = dmso_CD69_percent)+
geom_point(size = 2) +
facet_grid(Donor~Gate) + 
stat_summary(fun.y = "mean", geom = "line", aes(group = Targeting)) + 
labs(x = "ng/mL (log10)", y = "Percent in gate")
  
}


percent_plot2 <- function(df_to_plot){
  ggplot(df_to_plot, aes(x = log10(NgPerMl), y = Percent, color = Gate)) + 
geom_hline(aes(yintercept = Percent, color = Gate),
           data = dmso_CD69_percent)+
geom_point(size = 2) +
facet_grid(Donor~Targeting) + 
stat_summary(fun.y = "mean", geom = "line", aes(group = Gate)) + 
labs(x = "ng/mL (log10)", y = "Percent in gate")
}


```

### anti-CD4 antibody targeting and isotype control

```{r percent plot helper function}

cd4_cd4_iso_percent <- CD69_percent%>%
  filter(Targeting == "CD4 antibody"| Targeting == "Isotype for\nCD4 antibody" | Targeting == "Non-Targeted\nNanoparticles"|Targeting == "Free Drug")
  
  
#percent_plot(cd4_cd4_iso_percent)


percent_plot2(cd4_cd4_iso_percent)





```

### Fragment anti-CD4 antibody targeting and isotype control

```{r frament and iso}

fragment_and_fiso_percent <- CD69_percent%>%
  filter(Targeting == "Fragment\nCD4 antibody"| Targeting == "Isotype for\nCD4 fragment\nantibody"| Targeting == "Non-Targeted\nNanoparticles"|Targeting == "Free Drug")
  
  
  

#percent_plot(fragment_and_fiso_percent)



percent_plot2(fragment_and_fiso_percent)
```


### BP4 targeting

```{r bp4}

bp4_percent <-CD69_percent%>%
  filter(Targeting == "BP4"|Targeting == "Non-Targeted\nNanoparticles"|Targeting == "Free Drug")
  
#percent_plot(bp4_percent)

percent_plot2(bp4_percent)

```

## Median fluorescence intensity of CD69

Symbols represent experimental replicates and the connecting lines show the average of the replicates. The horizontal lines represent the DMSO negative controls.

```{r}
# To show the MFI in the same format as I did for percent CD69, I need to prepare the data in the same way, but use the values from the MFI columns in the "tests" df and the "dmso" df.



CD69_MFI <- tests %>%
  select(CD8CD69MFI, CD8NegCD69MFI, Formulation, Targeting, NgPerMl, Donor,Drug)%>%
  gather(data = ., key = Gate, value = MFI, CD8CD69MFI, CD8NegCD69MFI) 

#the values in the Targeting column are long and don't fit well on my plots so I'll add some line breaks here.



dmso_CD69_MFI <-dmso %>%
  select(CD8CD69MFI, CD8NegCD69MFI,Donor)%>%
  gather(data = ., key = Gate, value = MFI,CD8CD69MFI, CD8NegCD69MFI)



#here is a function for making the plots for the CD69 MFI results

MFI_plot <-function(df_to_plot){
  
ggplot(df_to_plot, aes(x = log10(NgPerMl), y = MFI, color = Targeting)) + 
geom_hline(aes(yintercept = MFI),
           data = dmso_CD69_MFI)+
geom_point(size = 2) +
facet_grid(Donor~Gate) + 
stat_summary(fun.y = "mean", geom = "line", aes(group = Targeting)) + 
labs(x = "ng/mL (log10)", y = "MFI")
  
}


#Here is a function for a different looking MFI plot

MFI_plot2 <- function(df_to_plot){
  ggplot(df_to_plot, aes(x = log10(NgPerMl), y = MFI, color = Gate)) + 
geom_hline(aes(yintercept = MFI, color = Gate),
           data = dmso_CD69_MFI)+
geom_point(size = 2) +
facet_grid(Donor~Targeting) + 
stat_summary(fun.y = "mean", geom = "line", aes(group = Gate)) + 
labs(x = "ng/mL (log10)", y = "MFI")
}

```




### anti-CD4 antibody targeting and isotype control


```{r CD4 and iso mfi}

cd4_cd4_iso_MFI <- CD69_MFI%>%
  filter(Targeting == "CD4 antibody"| Targeting == "Isotype for\nCD4 antibody" | Targeting == "Non-Targeted\nNanoparticles"|Targeting == "Free Drug")


#MFI_plot(cd4_cd4_iso_MFI)

MFI_plot2(cd4_cd4_iso_MFI)


```

### Fragment anti-CD4 antibody targeting and isotype control


```{r fragment and iso mfi}


fragment_and_fiso_MFI <- CD69_MFI%>%
  filter(Targeting == "Fragment\nCD4 antibody"| Targeting == "Isotype for\nCD4 fragment\nantibody"| Targeting == "Non-Targeted\nNanoparticles"|Targeting == "Free Drug")
  

#MFI_plot(fragment_iso_MFI)

MFI_plot2(fragment_and_fiso_MFI)
```

### BP4 targeting

```{r bp4 mfi}

bp4_MFI <-CD69_MFI%>%
  filter(Targeting == "BP4"|Targeting == "Non-Targeted\nNanoparticles"|Targeting == "Free Drug")
  
#MFI_plot(bp4_MFI)

MFI_plot2(bp4_MFI)

```

### MFI ratios

The following plot shows the ratio of the CD8- MFI over the CD8+ MFI. The horizontal line is at 1, where the MFI's are equal.

```{r MFI ratios}

tests <- tests %>%
  mutate(CD8Neg_CD8_MFI_ratio = CD8Neg/CD8)


tests$Targeting <- factor(tests$Targeting, levels = c("CD4 antibody", "Isotype for\nCD4 antibody", "Fragment\nCD4 antibody","Isotype for\nCD4 fragment\nantibody", "BP4","Free Drug", "Non-Targeted\nNanoparticles" ))


ggplot(tests, aes(x = log10(NgPerMl), CD8Neg_CD8_MFI_ratio )) +
  geom_point(aes())+
  stat_summary(fun.y = "mean", geom = "line", aes()) +
  facet_grid(Donor~Targeting)+
  labs(x = "ng/mL (log10)", y = "CD8-/CD8 MFI")+
  geom_hline(yintercept = 1, color = "blue")+
  theme(strip.text = element_text(size = 8))

```

## Conclusions

### CD4+ staining: problems with nanoparticles

It looks like the targeting nanoparticles interfere with the phenotyping antibody, so we should use CD8- going forward to represent CD4+ T cells.

### Percent CD69+: similar results within and between cell types

Within cell types: We saw a similar percent of cells in the CD69+ gate within cell types when we compared the nanoparticle formulations and their isotype controls or non-targeting nanoparticles. 

Between cells types: Across all formulations CD8- T cells had consistantly lower CD69+ percentages at higher drug concentrations, but the difference was quite small. In general, results do not show a strong targeting effect. 


### CD69 MFI: CD8- has higher MFI, but no obvious targeting effect

Within cell types: As with the %CD69 results, we saw similar responses between targeting formulations and their controls, with a small trend for slightly lower CD69 MFI's in the targeted formulations in CD8- cells.

Between cell types: CD69 MFI's were consistantly higher in CD8- cells, particularly in Donor B.


### Next steps

* Include HLA-DR, CD28 and CXCR4 in staining panel again to show inflammatory (or lack thereof) response.

* Try small steps in the serial dilutions. Here we did 3-fold steps, could go down to 2 or 1.5.

* Maybe try different incubation times/time course experiment?
