---
title: "Toxicity of CD4 targeting nanoparticles with chemically conjugated MPA"
output: html_document
---

## Goal
Determine toxicity of CD4 targeting nanoparticles with chemically conjugated MPA compared to isotype control, free MPA, non-targeted nanoparticles with MPA and non-targeted nanoparticles without MPA and DMSO.

## Experiment

4 and 7 day exposures of cytokine stimulated HIV- human PBMC from 2 donors to various nanoparticle formulations followed by phenotyping and viability measurements by flow cytometry.

## Conditions

Drug | Targeting | Formulation
----|---------|-----------
MPA|CD4|nanoparticle
MPA|CD4 iso|nanoparticle
no drug|CD4|nanoparticle
no drug|CD4 iso|nanoparticle
MPA|no targeting|nanoparticle
no drug|no targeting|nanoparticle
MPA|no targeting|free drug
DMSO|no targeting|free drug

## Concentrations

* Free MPA starting concentration: 400 ng/mL

* Nanoparticle formulations starting concentration: 8,000 ng/mL (20x free drug)

* For formulations without any MPA, we used the same concentration of nanoparticles as the conditions with MPA.


## Caveats

Resuspending the nanoparticles took about 10mins of sonicating. We started by sonciating 2mins, then vortex, then sonicate 3mins, but after that there were still visible aggregates at the bottom of the tubes. You don't see this if you look right after vortexing because they are still suspended, but if you wait a minute they are visible once they fall to the bottom of the tube. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plater)
library(stringr)
library(ggplot2)
```


```{r read in flow data}


day4 <- read.csv("J:/MacLabUsers/HLADIK SHARED/Projects/amFAR nanoparticles/003_MPA/003_MPA_Day_4.csv")


day7 <- read.csv("J:/MacLabUsers/HLADIK SHARED/Projects/amFAR nanoparticles/003_MPA/003_MPA_Day_7.csv")

```



```{r}
# read in plate layout data 
dir <- "J:/MacLabUsers/HLADIK SHARED/Projects/amFAR nanoparticles/003_MPA/plate_layout"



mapping_files <- paste0(dir, c("_1_5","_2_6", "_3_7", "_4_8"), ".csv")




#read in the plate layouts
plate_mappings_day4 <- plater::read_plates(mapping_files, well_ids_column = "Well", plate_names = c("plate_1","plate_2","plate_3","plate_4"))

plate_mappings_day7 <- plater::read_plates(mapping_files, well_ids_column = "Well", plate_names = c("plate_5","plate_6","plate_7","plate_8"))

```


```{r}


#combine the day4 data and the plate layout meta data, then make a "day" column and fill it wil "4"
day4 <- left_join(day4, plate_mappings_day4, by = c("Well","Plate"))
day4 <- mutate(day4, day = "4")

#same as above but for day7
day7 <- left_join(day7, plate_mappings_day7, by = c("Well","Plate"))
day7 <- mutate(day7, day = "7")

day_both <- rbind(day4,day7)



```





## Viability

Each symbol represents a single well (each condition was run in two wells)

Horizontal lines represent the average percentage of live cells in the DMSO-treated wells.

Lines between the dots represent the average among replicates.



```{r create DMSO and nonDMSO dfs}

dmso <-  day_both %>%
  filter(Drug == "DMSO")%>%
  group_by(day,Plate, Donor) %>% 
  summarise_at(vars(CD4:CD8NegCount), funs(mean))%>%
  ungroup()

# get just the non-control wells, and exclude the "NP only" wells, which just had nanoparticles (no cells) and were used for gating.

no_DMSO<-filter(day_both, Drug != "DMSO",Targeting != "NP only")

```



```{r plot viability}

#all data together

# ggplot(no_DMSO, aes(x = log10(MPA_ng_per_mL), y = Live, color = Donor, shape = day)) + 
#     geom_hline(data = dmso, aes(yintercept = Live, color = Donor)) +
#     geom_point(size = 2) +
#     facet_grid(Drug~Targeting+Formulation) + 
#     stat_summary(fun.y = "mean", geom = "line", aes(group = interaction(Donor, day))) + 
#     xlab("MPA Concentration (ng/mL)")
```

### Targeting and non-targeting nanoparticles *without* MPA

The x-axis shows *nanoparticle* concentration.

```{r targeting effects of NP only}

targeting_only <- no_DMSO %>%
  filter(Drug == "no drug")

ggplot(targeting_only, aes(x = NP_mg_per_mL, y = Live, color = Donor)) + 
  scale_x_log10()+
    geom_hline(data = dmso, aes(yintercept = Live, color = Donor)) +
    geom_point(size = 2) +
    facet_grid(day~Targeting+Formulation) + 
    xlab("Nanoparticle Concentration mg/mL (log10 scale)")+
  stat_summary(fun.y = "mean", geom = "line", aes(group = Donor))

```

### Targeting and non-targeting nanoparticles *with* MPA and Free Drug

The x-axis shows *MPA* concentration

```{r Effect of Free MPA vs untargetd NP with MPA}


MPA_cc_or_free_drug <- no_DMSO %>%
filter(Drug == "MPA_cc")


ggplot(MPA_cc_or_free_drug, aes(x = log10(MPA_ng_per_mL), y = Live, color = Donor)) + 
    geom_hline(data = dmso, aes(yintercept = Live, color = Donor)) +
    geom_point(size = 2) +
    facet_grid(day~Targeting+Formulation) + 
    labs(x = "MPA Concentration ng/mL (log10 transformed)", y = "Percent Live")+
  stat_summary(fun.y = "mean", geom = "line", aes(group = Donor))


```


```{r alternate plot arrangement}

ggplot(MPA_cc_or_free_drug, aes(x = log10(MPA_ng_per_mL), y = Live, color = day)) + 
    geom_hline(data = dmso, aes(yintercept = Live, color = day)) +
    geom_point(size = 2) +
    facet_grid(Donor~Targeting+Formulation) + 
    labs(x = "MPA Concentration ng/mL (log10 transformed)", y = "Percent Live")+
  stat_summary(fun.y = "mean", geom = "line", aes(group = day))

```



```{r }

### Comparing cells with and without MPA on day 4 and 7.


# Plate_1_2_5_6 <- no_DMSO %>%
#   filter(Plate == "plate_1" | Plate == "plate_2" | Plate == "plate_5"| Plate == "plate_6")
# 
# 
# ggplot(Plate_1_2_5_6, aes(x = NP_mg_per_mL, y = Live, color = Donor)) + 
#     geom_hline(data = dmso, aes(yintercept = Live, color = Donor)) +
#     geom_point(size = 2) +
#     facet_grid(day~Targeting + Drug) + 
#     xlab("Nanoparticle Concentration (mg/mL)") +
#     stat_summary(fun.y = "mean", geom = "line", aes(group = Donor)) + 
#     scale_x_log10()

```


```{r }
#want to keep data where either CD8Count >100 or CD8Neg>100, otherwise remove the row.

#To get CD8 high count data, select all cols EXCEPT CD8Neg ones. Filter for CD8Count >100. Gather. Repeat for CD8Neg, then rbind.

CD8_high_count <- no_DMSO %>%
  select(-CD8Neg, -CD8NegCount)%>%
  filter(CD8Count > 100)%>%
  gather(., key = Gate , value = Measurement,CD4:CD8Count)
  

CD8Neg_high_count <- no_DMSO %>%
  select(-CD8Count, -CD8)%>%
  filter(CD8NegCount > 100)%>%
  gather(., key = Gate , value = Measurement,CD4:CD8NegCount)


high_count <- rbind( CD8_high_count, CD8Neg_high_count)
```

### Day 4: Percentage in gate

Data not shown where CD8+ OR CD8- counts were <100 cells 
```{r }

high_count %>%
  filter (day == "4") %>%
  filter(Gate == "CD8"|Gate == "CD8Neg" | Gate == "CD4")%>%
ggplot(., aes(x = log10(MPA_ng_per_mL), Measurement )) +
  geom_point(aes(color = Gate))+
  stat_summary(fun.y = "mean", geom = "line", aes(group = Gate, color = Gate)) +
  facet_grid(Donor~Targeting)+
  labs(x = " MPA Concentration ng/mL (log10 transformed)", y = "Percent in gate")

```


### Day 7: Percentage in gate

Data not shown where CD8+ OR CD8- counts were <100 cells 

```{r }

high_count %>%
  filter (day == "7") %>%
  filter(Gate == "CD8"|Gate == "CD8Neg" | Gate == "CD4")%>%
ggplot(., aes(x = log10(MPA_ng_per_mL), Measurement )) +
  geom_point(aes(color = Gate))+
  stat_summary(fun.y = "mean", geom = "line", aes(group = Gate, color = Gate)) +
  facet_grid(Donor~Targeting)+
  labs(x = "MPA Concentration ng/mL (log10)", y = "Percent in gate")+
  theme(strip.text = element_text(size = 8))

```

 



```{r }

# ### Day 4: Percentage in gate
# #Data not shown where CD8+ OR CD8- counts were <100 cells 


# high_count %>%
#   filter (day == "4") %>%
#   filter(Gate == "CD8"|Gate == "CD8Neg" | Gate == "CD4")%>%
# ggplot(., aes(x = NP_mg_per_mL, Measurement )) +
#   geom_point(aes(color = Gate))+
#   stat_summary(fun.y = "mean", geom = "line", aes(group = Gate, color = Gate)) +
#   facet_grid(Donor~Targeting)+
#   labs(x = " Nanoparticle Concentration, mg/mL(log10)", y = "Percent in gate")+
#   scale_x_log10()

```

 

```{r }
### Day 7: Percentage in gate

#Data not shown where CD8+ OR CD8- counts were <100 cells


# high_count %>%
#   filter (day == "7") %>%
#   filter(Gate == "CD8"|Gate == "CD8Neg" | Gate == "CD4")%>%
# ggplot(., aes(x = NP_mg_per_mL, Measurement )) +
#   geom_point(aes(color = Gate))+
#   stat_summary(fun.y = "mean", geom = "line", aes(group = Gate, color = Gate)) +
#   facet_grid(Donor~Targeting)+
#   labs(x = "Nanoparticle Concentration, mg/mL (log10)", y = "Percent in gate") + 
#   scale_x_log10()

```


### Conclusions

1) Donor B cells have more viability than Donor A cells and this effect is more pronounced in the nanoparticle treated cells.

2) Higher concentration of Nanoparticals are affecting the viability of cells adversely.

3) Addition of Drug to nanoparticles seems to increase the viability a bit especially in 3rd to 5th dilution.

4) There is not much pronounced difference between the viability of cells between day 4 and day 7.

4) Viability in cells treated with CD4_iso is more than cells treated with CD4.

5) CD4+ staining does not work when you use CD4-targeting nanoparticles
