---
title: "Proliferation experiments analysis"
author: "Claire Levy"
output: github_document
---



## Proliferation assay 001: MPA

### Goal
Assess anti-proliferative effects of 3 different MPA treatments (free drug and two nanoparticle formulations) in human PBMC from two HIV - donors, using two activation methods (beads and cytokines). 

### Treatments

1. Free MPA
2. MPA loaded LCNP (physically encapsulated)
3. MPA conjugated LCNP (conjugated with PLGA)

### Stimulations

All cells were treated with IL2 and one of the following stimulation cocktails:

Beads: CD3/CD28 beads (5 days, “antigenic”)

Cytokines:IL15,IL7 (7 days, “homeostatic”)




### Doses

We did titrations of MPA in 2-fold steps:
 
Free MPA: 1.6-0.00625 $\mu$g/mL and 0
	
MPA loaded LCNP: 1.6-0.00625 $\mu$g/mL and 0
	
MPA conjugated LCNP, use 50x more than others, 80-0.3125 $\mu$g/mL and 0

### Replicates

All treatments were done with two replicates per donor.

### Caveats

It was somewhat difficult to completely resuspend the nanoparticle formulations in the media. We vortexed and sonicated but for both formulations were left with some aggregates that we couldn't get rid of. 

We ran out of the Ab that we were using for CD4 staining so we had to use less that the amount called for in the protocol. The results looked ok though and we were able to see clear CD4+ populations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plater)
library(stringr)

```


```{r read in plate layouts metadata}

layouts001_dir<-"../001 proliferation MPA/plate_layouts"

layouts001_files <-paste0(layouts001_dir,"/", list.files(layouts001_dir))


#giving a plate name with same values as those in the PLATE NAME column in the flowjo files.
layouts001<- read_plates(layouts001_files, plate_names = c("IL2_CD3_CD28_1801", "IL2_CD3_CD28_5140","IL2_IL7_IL15_1801", "IL2_IL7_IL15_5140"))


names(layouts001)[1]<-"PLATE NAME"
```


```{r read in flowjo data}

#read in data tables from flowjo: There is one for the bead stimulated samples and one for the cytokine stimulated ones for each donor. 

data001_dir<-"../001 proliferation MPA/data/flowjo_tables"

data001_files <-paste0(data001_dir,"/", list.files(data001_dir))

#read in as a list of dfs
data001<-lapply(data001_files, FUN = read.csv)

#bind into one df
data001<-bind_rows(data001)


#clean up the col names so it's not terrible, the % of parent data are first, then the count data

colnames(data001)<-c("Wells","PLATE NAME","live","CD14_neg","CD3","CD4","CD4_proliferated","CD8","CD8_proliferated", "live_count","CD14_neg_count","CD3_count","CD4_count","CD4_proliferated_count","CD8_count","CD8_proliferated_count", "X.1")




#The well column is messed up, I only want the first 3 characters and I don't want the empty last column ("X.1") AND I don't want the rows with entries for "Mean" and "SD" in the Wells column
     
data001<-data001 %>%
  mutate(Wells = substr(Wells, 1,3))%>%
  select(-X.1)%>%
  filter(Wells != "SD" & Wells != "Mean" & Wells != "Mea")



```

```{r merge metadata and flowjo data}




merge_data001<- merge(layouts001, data001, by = c("Wells", "PLATE NAME"))

#make it more clear that the conjugated MPA is 50X more concentrated than the other ones:

merge_data001<-merge_data001 %>%
mutate(Formulation = str_replace(Formulation,"MPA conj. NP", "50X MPA conj. NP"))%>%
  mutate(Stimulation = str_replace(Stimulation, "IL2_IL7_IL15", "Cytokines"))%>%
  mutate(Stimulation = str_replace(Stimulation,"IL2_CD3_CD28", "Beads"))



merge_data001$Formulation<-factor(merge_data001$Formulation, levels = c("Free drug", "MPA loaded NP", "50X MPA conj. NP"))


#make a column for if there were <100 CD4 or CD8 cells and <1000 live cells


merge_data001<- merge_data001 %>%
  mutate("CD4_less_than_100" = ifelse(CD4_count <100, "yes", "no"))%>%
  mutate("CD8_less_than_100" = ifelse(CD8_count <100, "yes", "no"))%>%
  mutate("live_less_than_1k" = ifelse(live_count <1000, "yes", "no"))

```
## Results

### Viability

In general, viability decreases as the MPA concentration increases. The decrease is somewhat more dramatic for the cytokine stimulated cells. 

The black line represents the average viability between the two donors and the shape indicates if there were fewer than 1,000 cells in the live gate.

```{r plot viability}

ggplot(merge_data001,aes(x = ug_per_mL_MPA, y = live))+
  geom_point(aes(color = Donor, shape = live_less_than_1k), alpha = 0.5, size = 2.5)+
  facet_grid(Stimulation~Formulation, scales = "free_x")+
  stat_summary(fun.y = mean, geom = "line")+
  ggtitle("Percent viability: all cells")+
  scale_x_log10()+
  labs(y = "Percent Live", x = "\U03BCg/mL MPA")+
  theme(axis.title.y = element_text(size =16))
```

### Proliferation: CD4+ and CD8+ T cells

Proliferation started out higher for the CD3/CD28 bead stimulated cells and there was also more variation between the two donors for this condition.

Proliferation dropped off at the same point in the titration for all conditions except for 50X MPA conj. NP with cytokine stim, where the pattern is different.

The shapes indicate if there were less than 100 cells in the CD4+ or CD8+ gate.

```{r plot percent proliferated CD4}
ggplot(merge_data001,aes(x = ug_per_mL_MPA, y = CD4_proliferated))+
  geom_point(aes(color = Donor, shape = CD4_less_than_100), alpha = 0.5, size = 2.5)+
  facet_grid(Stimulation~Formulation, scales = "free_x")+
  stat_summary(fun.y = mean, geom = "line")+
  ggtitle("Percent proliferation of CD4+ cells")+
  scale_x_log10()+
  labs(y = "Percent Proliferated", x = "\U03BCg/mL MPA")+
  theme(axis.title.y = element_text(size =16))
```


```{r plot percent proliferated CD8}
ggplot(merge_data001,aes(x = ug_per_mL_MPA, y = CD8_proliferated))+
  geom_point(aes(color = Donor, shape = CD8_less_than_100), alpha = 0.5, size = 2.5)+
  facet_grid(Stimulation~Formulation, scales = "free_x")+
  stat_summary(fun.y = mean, geom = "line")+
  ggtitle("Percent proliferation of CD8+ cells")+
  scale_x_log10()+
  labs(y = "Percent Proliferated", x = "\U03BCg/mL MPA")+
  theme(axis.title.y = element_text(size =16))
```

