---
title: "Proliferation experiments analysis"
author: "Claire Levy"
output: html_document
---



## Proliferation assay 001: MPA

### Goal
Assess anti-proliferative effects of 3 different MPA treatments (free drug and two nanoparticle formulations) in human PBMC (two donors) using two activation methods (beads and cytokines). 

### Treatments

1. Free MPA
2. MPA loaded LCNP (physically encapsulated)
3. MPA conjugated LCNP (conjugated with PLGA)

### Stimulations

IL2 and CD3/CD28 beads (5 days, “antigenic”)
IL2,IL15,IL7 (7 days, “homeostatic”)


### Doses
 We did titrations in 2- fold steps
	Free MPA: 1.6-0.00625 ug/mL and 0
	MPA loaded LCNP: 1.6-0.00625 ug/mL and 0
	MPA conjugated LCNP, use 50x more than others, 80-0.3125 ug/mL and 0



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(tidyverse)
library(plater)
library(stringr)

```


```{r read in plate layouts metadata}

layouts001_dir<-"../001 proliferation MPA/plate_layouts"

layouts001_files <-paste0(layouts001_dir,"/", list.files(layouts001_dir))


#giving a plate name with same values as those in the PLATE NAME column in the flowjo files.
layouts001<- read_plates(layouts001_files, plate_names = c("IL2_CD3_CD28_1801", "IL2_CD3_CD28_5140","IL2_IL7_IL15_1801", "IL2_IL7_IL15_5140"))


names(layouts001)[1]<-"PLATE NAME"
```


```{r read in flowjo data}

#read in data tables from flowjo: There is one for the bead stimulated samples and one for the cytokine stimulated ones for each donor. 

data001_dir<-"../001 proliferation MPA/data/flowjo_tables"

data001_files <-paste0(data001_dir,"/", list.files(data001_dir))

#read in as a list of dfs
data001<-lapply(data001_files, FUN = read.csv)

#bind into one df
data001<-bind_rows(data001)


#clean up the col names so it's not terrible

colnames(data001)<-c("Wells","live","CD14_neg","CD3","CD4","CD4_proliferated","CD8","CD8_proliferated","PLATE NAME","X.1")


#The well column is messed up, I only want the first 3 characters and I don't want the empty last column ("X.1") AND I don't want the rows with entries for "Mean" and "SD" in the Wells column
     
data001<-data001 %>%
  mutate(Wells = substr(Wells, 1,3))%>%
  select(-X.1)%>%
  filter(Wells != "SD" & Wells != "Mean" & Wells != "Mea")

```

```{r merge metadata and flowjo data}
merge_data001<- merge(layouts001, data001, by = c("Wells", "PLATE NAME"))
```

```{r plot viability}

ggplot(merge_data001,aes(x = ug_per_mL_MPA, y = live))+
  geom_point(aes(),color = "lightblue")+
  facet_grid(Stimulation~Formulation, scales = "free_x")+
  stat_summary(fun.y = mean, geom = "line")
```


Big difference between donors here for the cytokine stimulated condition... check the gates here.

```{r plot percent proliferated CD4}
ggplot(merge_data001,aes(x = ug_per_mL_MPA, y = CD4_proliferated))+
  geom_point(aes(color=Donor), alpha = 0.5)+
  facet_grid(Stimulation~Formulation, scales = "free_x")+
  stat_summary(fun.y = mean, geom = "line")
```

Donors similar here.

```{r plot percent proliferated CD8}
ggplot(merge_data001,aes(x = ug_per_mL_MPA, y = CD8_proliferated))+
  geom_point(aes(color=Donor), alpha = 0.5)+
  facet_grid(Stimulation~Formulation, scales = "free_x")+
  stat_summary(fun.y = mean, geom = "line")
```