---
output: word_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)




experiment<-"LRA_003"

#MPA molecular weight = 320.341 g/mol
#PMA molecular weight = 616.83 g/mol

#molarity = mol/L
#uM/L x 1mol/1E6umol x g/mol x 1E6ug/g x 1L/1000mL
#units cancel, leaving ug/mL
#the two 1E6's cancel, so you just need to multiply by mw and divide by 1000

uM_to_ug_per_mL<- function(mw, uM){
  ug_per_mL<-(uM * mw)/1000
  round(ug_per_mL, digits = 2)
}

#ug/mL x 1g/1E6ug x mol/g x 1000mL/L x 1e6umol/mol
#units cancel, leaving umol/L, which = uM
#The two 1E6 cancel, so you just have to divide by mw (multiply by inverse) and multiply by 1000.

ug_per_mL_to_uM<-function(mw, ug_per_mL){
 uM<-(ug_per_mL/mw) *1000
 round(uM, digits = 2)
}
 

#enter concentrations as x/mL where x is the same units. Result is uL
how_much_to_use<- function(c1, c2, v2){
  v1<-(c2 *v2)/c1
  v1uL<-v1 *1000
  round(v1uL, digits = 4)
}

```

## `r experiment` Protocol

**Goal**

Compare targeting of various formulations of ingenol nanoparticles (conjugated, not encapsulated) by measuring CD69-expression on CD8 and CD4 T cells after 24h.

**Treatments: To be done in duplicate**

1. Free ingenol

2. Ingenol nanoparticles

3. Ingenol nanoparticles with BP4 (CD4-targeting peptide)

4. Ingenol nanoparticles with CD4 antibody

5. Ingenol nanoparticles with isotype antibody

6. Ingenol nanoparticles with fragment CD4 antibody

7. Ingenol nanoparticles with fragment isotype antibody

8. DMSO (ingenol vehicle)
  
**Number of wells**

```{r}
DONORS <- 2

TREATMENTS <- 7

CONCENTRATIONS <- 6

REPLICATES <- 2

EXPERIMENTAL_WELLS <- DONORS * TREATMENTS * CONCENTRATIONS * REPLICATES

CONTROL_WELLS <- DONORS * REPLICATES

NUMBER_OF_WELLS <- EXPERIMENTAL_WELLS + CONTROL_WELLS
```

For the non-control conditions, we need `r DONORS` donors x `r TREATMENTS` treatments x `r CONCENTRATIONS` concentrations x `r REPLICATES` replicates = `r EXPERIMENTAL_WELLS` wells, `r DONORS * CONCENTRATIONS * REPLICATES`  per condition. 

For the control, we need `r DONORS` donors x `r REPLICATES` replicate = `r CONTROL_WELLS` wells

`r EXPERIMENTAL_WELLS` + `r CONTROL_WELLS` = `r NUMBER_OF_WELLS` wells total

**Donors**
`r DONORS` x HIV- , total PBMC 
150k cells per well x `r NUMBER_OF_WELLS` wells = `r 1.5E5 * NUMBER_OF_WELLS / DONORS` needed.

**Timing**

Treat with LRAs for 24hr, then stain cells for flow.

**Prepare plate for doing drug dilutions**

In each well, add 100uL R10. Later, 100uL of cells will be added to these wells, diluting the drug that is in there 2-fold, which is why we need to add the drug at a 2X concentration.


**Prepare drug treatments in tubes**

For each serial dilution (3-fold) need to start the dilution with 50uL of the highest concentration (50uL plus 100uL = 3-fold dilution). 

There will be 2 reps of each dilution x 2 donors so need to make at least 50uL x 4 = 150uL of each starting drug solution for the single drug conditions. 

Make 300uL to minimize tears.

For the controls, need 2 reps per control, 100uL each so make 450uL of the starting concentration for each (DMSO).

**Starting Drug Concentrations**
Drugs need to be added at 6X the final desired concentration, so that when we add 50uL to the 100uL already in the well, they will be diluted 1:3 to 2X, in preparation for being further diluted 1:2 to 1X when cells are added

**Treatment Doses**

We'll start with 9-fold over what SC used for the combination dose

  Free drug: Start with 9-fold over 1.5ng/mL = 13.5ng/mL = 0.0135ug/mL
  
  LCNP: Start with 9-fold over 24ng/mL = 216ng/mL = 0.216ug/mL
  
```{r}

mL_OF_6X_DRUG <- 0.3
uL_OF_6X_DRUG <- mL_OF_6X_DRUG * 1000

mL_OF_2X_CNTL<-.45

uL_OF_2X_CNTL<-mL_OF_2X_CNTL * 1000

```

**DMSO**

Use the highest concentration of any of the free drugs/3 (to make it 2X) and bring to `r mL_OF_2X_CNTL`mL


```{r ing3 concentrtaions, message=TRUE}
#Ing starting concentration for free drug single = 0.0135ug/mL
#Need 6X, 0.081ug/mL

Ing_FD_single<-how_much_to_use(500,0.081, mL_OF_6X_DRUG)

#Ing starting concentration for LCNP single = 0.216ug/mL
#Need 6X, 1.296ug/mL

Ing_6x_NP <- 1.296

Ing_NP_ug<-25.6
Ing_NP_mL<-1


Ing_NP_single<-how_much_to_use(Ing_NP_ug/Ing_NP_mL, Ing_6x_NP, mL_OF_6X_DRUG)

```

**6X Ing free drug single**

SC gave us Ing free drug at 500ug/mL

We only need a little amount, so make a 1:100 dilution of Ing_FD and then:

Use `r Ing_FD_single * 100` uL, bring to `r mL_OF_6X_DRUG` mL with R10 by adding `r mL_OF_6X_DRUG * 1000 - (Ing_FD_single * 100)` uL

Make enough 1:100 to use again for the combo treatment.


**6X Ing NP (chemically conjugated)**

Shijie brought us nanoparticles corresponding to 25.6 ug of ingenol for each formulation. He said that we should resuspend them in 1 mL R10 to `r Ing_NP_ug/Ing_NP_mL`ug/mL. Vortex and sonicate until there are no clumps. The below procedure will need to be repeated for each formulation.

Use `r Ing_NP_single` uL, bring to `r mL_OF_6X_DRUG`mL with R10 by adding `r mL_OF_6X_DRUG * 1000 - Ing_NP_single`uL. This gives `r Ing_6x_NP` ug/mL, which is six times (allows for dilution) our desired concentration of 0.216ug/mL for the highest concentration. 

###Add drugs and controls to plate
Add 50uL of 6X drug cocktails to starting well and do serial dilutions down to row G.

For controls, add 100uL of 2X.

###Add cells to plate
Donor A is VEU5031 and Donor B is VEU5099.

Dilute cells to 1.5E5/100uL = 1.5E6/mL. Add 100uL to each well.

###Incubate plates at 37C for 24hr

###Stain for flow

The flow staining protocol has three steps: 

* Live/dead (in PBS, at RT)
* CXCR4 (in FACS Wash, at 37C)
* Everything else (in FACS Wash, at RT)

Live/Dead: 

1. Prepare staining reagent: `r 0.1*NUMBER_OF_WELLS` µL Live/Dead Aqua + `r 49.9*NUMBER_OF_WELLS` µL PBS.
1. Spin cells at 750g for 3 minutes.
1. Resuspend cells in PBS.
1. Spin cells at 750g for 3 minutes. 
1. Resuspend cells in 45 µL staining reagent (this accounts for overage because we prepared 50 µL per well).
1. Incubate in the dark at RT for 20 min.
1. Add 150 µL FACS Wash to each well. 

Antibody staining: 

1. Prepare staining cocktail: 

* `r NUMBER_OF_WELLS * 5` µL CD8 PerCP-Cy5.5
* `r NUMBER_OF_WELLS * 5` µL CD3 v450
* `r NUMBER_OF_WELLS * 2.5` µL CD69 APC		
* `r NUMBER_OF_WELLS * 0.5` µL CD4 ECD
* `r NUMBER_OF_WELLS * 37` µL FACS Wash.

1. Spin cells at 750g for 3 minutes.
1. Resuspend cells in 45 µL staining cocktail (this accounts for overage). 
1. Incubate in the dark at RT for 20 min.
1. Add 150 µL FACS Wash to each well. 
1. Spin cells at 750g for 3 minutes.
1. Resuspend in 200 µL.
1. Spin cells at 750g for 3 minutes.
1. Resuspend in 100 µL 1% PFA in FACS Wash (prepare `r NUMBER_OF_WELLS * 0.1 * 1.1` mL by adding `r NUMBER_OF_WELLS * 0.01 * 1.1` mL 10% PFA to `r NUMBER_OF_WELLS * 0.09 * 1.1` mL FACS Wash)
1. Place at 4C. 