---
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)




experiment<-"LRA_001"



#MPA molecular weight = 320.341 g/mol
#molarity = mol/L
uM_to_ug_per_mL<- function(mw, uM){
  molsPerL<-uM * 1000
  molsPerMl<-(molsPerL * mw)/1e6
  round(molsPerMl, digits = 2)
}

uM_to_ug_per_mL(709.018,2)

ug_per_mL_to_uM<-function(mw, ug_per_mL){
 mols<-(ug_per_mL * 1e6)/mw
 uM<-mols/1000
 round(uM, digits = 2)
}
 

#enter concentrations as x/mL where x is the same units. Result is uL
how_much_to_use<- function(c1, c2, v2){
  v1<-(c2 *v2)/c1
  v1uL<-v1 *1000
  round(v1uL, digits = 2)
}




```

## `r experiment` Protocol



**Goal**
Preliminary assay with 3 donors to see how many CD4+ T cells we will need for assessing latency reversing ability of nanoparticle bound LRA’s via non-readthrough PCR on RNA from HIV+ ART+ CD4+ T cells. Based on methods described in Bullen et al,  doi:10.1038/nm.3489

**Treatments: To be done in duplicate**

1. PMA/Ionomycin (+ cntl)
2. DMSO (untreated)
3. Bryostatin (non NP, to see if we can replicate Bullen et al.)

**Donors**
Donors are HIV+ on ART with <40 copies/mL from the SAC samples. Bullen et al did <50 copies/mL for at least 6 mos.




**T cell purification**: 

Enrich for CD4+CD69-CD25-HLADR- using StemCell kit (cat# 17962, EasySep Human Resting CD4 T Cell Isolation kit)

1.	Thaw cells
2.	Count
3.	Save aliquot of each for flow
4.	Transfer into FACS tubes and centrifuge 500g for 5 min. 
5.	Resuspended in EasySep buffer at 50 million cells/mL. 
6.	Add 50 µL isolation cocktail per mL of sample.
7.	Add 10 µL CD25 depletion cocktail per mL of sample.
a.	Note that this is five times less than recommend, but improves recovery as indicated in the notes and tips section.  
8.	Incubate 10 min
9.	Cortex RapidSpheres for 30s and then add 75 µL per mL of sample. 
10.	Incubate 5 min.
11.	Bring up to 2.5 mL with FACS wash
12.	Put in magnet for 5 min
13.	Pour off, saving supernatant = resting CD4 T cells
14.	Place the new tube containing the poured off supernatant into the magnet 
15.	Incubate in the magnet for 5 min
16.	Pour off, saving supernatant = resting CD4 T cells
17.	Count
18.	Save aliquot for flow

 Decide which two cell amounts to test:

x:

y:       


*Bullen et al did this:*

*Enrich CD4+ by negative depletion :*
*CD4+ T cell Isolation Kit, Miltenyi Biotec*

*Then enrich for resting Tcells:*

*CD69 MicroBead Kit II, Miltenyi Biotec*
*CD25 MicroBeads, Miltenyi Biotec*
*Anti–HLA-DR MicroBeads, Miltenyi Biotec*



**T-cell culture**

Grow with 10uM T20 (fusion inhibitor) to prevent new infections.


**Treatment Doses**

PMA/Ionomycin: 50ng/mL PMA and 1uM Ionomycin
Bryostatin-1: 10nM

**Timing**

Treat with LRAs for 18hr (Bullen et al did both 6hr and 18hr and saw similar results), then harvest cells and isolate total RNA.


**Thaw/enrich/count**

Thaw 5 vials of PBMC per donor in R10 and see how many CD4+ resting T cells we can get using  the StemCell kit, to enrich for CD4+CD69-CD25-HLADR- . Then count and decide which two cell amounts to test.


Save some cells to check purity by flow.

Make two tubes of cells of 2 concentrations, one that will give you X cells in 100uL and one that will give you Y cells in 100uL.
In a 96 well round bottom plate, add 100uL of cells to the wells as in the plate layout. Will add 100uL of media + drug to bring to a total of 200uL per well.


**Prepare drug treatments in tubes**


```{r}

total_wells <- function (donors , treatments, stimulations, concentrations, replicates){
  
  donors * treatments * stimulations * concentrations*replicates
}

cell_vol<-100

drug_vol<-100



number_drug_preparations <-3

```


We need `r ((total_wells(3, 3,1,2,2)* drug_vol)/number_drug_preparations)`uL per drug preparation and the solution will be diluted 1:2 with the cells. Make
1.5mL to have extra.


```{r}

mL_OF_2X_DRUG <- 1.5
uL_OF_2X_DRUG <- mL_OF_2X_DRUG * 1000

PMA<-how_much_to_use(1, .1, mL_OF_2X_DRUG)
Iono<- how_much_to_use(35,uM_to_ug_per_mL(709.018,2), mL_OF_2X_DRUG)
T20<-round(uL_OF_2X_DRUG / 22.5, digits = 2)

Bryo<-how_much_to_use(2000, 20, mL_OF_2X_DRUG)
DMSO<-uL_OF_2X_DRUG/100

```

**PMA/Ionomycin**

PMA: stock = 1ug/mL, want .1ug/mL, 2x of the final concentration.

Use `r PMA`uL

Ionomycin: stock = 35ug/mL, want 2uM,  2x of final concentration.

Use `r Iono`uL

T20: stock is at 45x and we need it at 2x so dilute 1:22.5

Use: `r T20`uL

R10: Bring to 1.5mL with R10:

Use `r uL_OF_2X_DRUG - (PMA + Iono +T20) `uL

**Bryostatin-1**

Bryostatin-1: stock is 200x and want 2000nM, 2x of final concentration

Use `r Bryo` uL

T20: stock is at 45x and we need it at 2x so dilute 1:22.5

Use: `r T20`uL

R10: Bring to 1.5mL with R10:

Use `r uL_OF_2X_DRUG - (T20 + Bryo)`uL

**DMSO**

To match Bryostatin, do a 1:100 dilution

Use: `r DMSO`uL

T20: stock is at 45x and we need it at 2x so dilute 1:22.5

Use: `r T20`uL

R10: Bring to 1.5mL with R10:

Use `r uL_OF_2X_DRUG - (T20 + DMSO)`uL
